A White Noise Approach to Evolutionary Ecology

### 2020-04-28 Current Version/Untitled.Rmd
reorganize

---
title: "A White Noise Approach to Evolutionary Ecology"
author: "Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon"
date: "3/1/2020"
abstract: "We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued stochastic processes. This approach allows us to compute the full implications of demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a quantitative genetic model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests absolute values of linear selection gradients are negatively correlated with competition coefficients when the ratio of interspecific trait variation to intraspecific trait variation is approximately less than $1.25$ and positively correlated otherwise. The inverse relationship holds for quadratic selection gradients except with threshold diversity ratio equal to unity instead of $1.25$. Hence, this approach contributes to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of stochastic models that underlie rigorous investigations of the relationship between feedbacks of biological processes and the patterns of diversity they produce."
bibliography: ref.bib
output:
  pdf_document: 
    extra_dependencies:
      - float
      - flafter
    fig_caption: yes
    fig_height: 3
    fig_width: 4
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{mathtools}
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
editor_options: 
  chunk_output_type: console
---

@@

# Introduction

@@

# The framework

@@

## Deterministic dynamics \label{deterministic}

Our review begins by considering the dynamics of an asexually reproducing population in a homogeneous environment. For simplicity, we first assume individuals are haploid and carry one of $K$ alleles each with a different fitness expressed as growth rate before introducing a model involving a quantitative trait. Under these assumptions, the derivation of the evolution of allele frequencies due to natural selection can be derived from expressions of exponential growth. This, and a few related approaches, have been provided by Crow and Kimura [-@9781932846126, \S 5.3]. Specifically, denoting $\nu_i$ the abundance of individuals with allele $i$ and $m_i$ the growth rate of allele $i$ [called the Malthusian parameter in @9781932846126], we have
\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation}
Starting from this model, we get the total abundance of the population as $N=\sum_{i=1}^K\nu_i$, the frequency of allele $i$ as $p_i=\nu_i/N$ and the mean fitness of the population as $\bar m=\sum_{i=1}^Kp_im_i$. Hence, we can employ some elementary calculus to derive the dynamics of abundance $dN/dt$ and the dynamics of allele frequencies $dp_1/dt,\dots,dp_K/dt$ as
\begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation}
\begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation}
Two important observations of these equations include (i) mean fitness $\bar m$ determines the abundance dynamics of the entire population and (ii) allele $i$ will increase (decrease) in frequency if $m_i>\bar m$ ($<\bar m$). Equation (\ref{popgen}) is known in the field of evolutionary game theory as the replicator equation [@Hofbauer1998; @9780674023383; @Lion2018; @Taylor1978; @Schuster1983]. Instead of being explicitly focused on alleles, the replicator equation describes the fluctuations of relative abundances of various _types_ in a population in terms of the vital rates of each type. Using a matrix of transition rates between differing types, it is straight-forward to extend the replicator equation to include mutation, which is known as the replicator-mutator equation [@9780674023383].

@@

To stay within the realm of biological plausibility we require a set technical assumptions. First, we assume the initial abundance density is continuous, non-negative, integrable and has finite trait mean and variance. That is, we assume $\nu(x,0)$ is continuous in $x$, satisfies $\nu(x,0)\geq0$ for all $x\in\mathbb{R}$ and
\begin{equation}
N(0)=\int_{-\infty}^{+\infty}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{-\infty}^{+\infty}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{-\infty}^{+\infty}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}
where $\bar x(t)$ and $\sigma^2(t)$ are respectively the mean trait and phenotypic variance at time $t\geq0$. Second, we assume selection is determined by the growth rate $m(h,x)$ that is differentiable with respect to both arguments and is bounded above by some $r\in\mathbb{R}$ for all $x\in\mathbb{R}$ and $h\geq0$. We also assume $(1+h)C\leq m(h,x)$ for some constant $C>0$ so that $m(h,x)$ is linear or more than linear in $h$. We denote by $K$ an operator that takes functions such as $\nu(x,t)$ as its argument and returns other functions that may still depend on the spatial variable $x$ and the temporal variable $t$. For our application we will be focused on operators $K$ that can be written as $(K\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x-y)\nu(y,t)dy$ for some non-negative and bounded function $\kappa$. We set $h=(K\nu)(x,t)$ to account for nonlocal effects of the abundance density $\nu(x,t)$ on the fitness of individuals with trait value $x$ [@Volpert2014]. As an example of a nonlocal effect, consider a population experiencing intraspecific competition for a common resource such that the competitive interactions are not mediated by any set of traits. In this case all individuals in the population effect the fitness of one another even though they may exhibit a diverse range of trait values. The term nonlocal comes from interpreting the trait value as spatial location. From here on we abbreviate $m(h,x)=m\big((K\nu)(x,t),x\big)$ to just $m(\nu,x)$. Third, we assume mutation is captured by diffusion with coefficient $\frac{\mu}{2}$. With these technicalities addressed we model the demographic dynamics of a population and the dynamics of a quantitative character simultaneously by the PDE
\begin{equation}\label{eq1}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)
\end{equation}
with the initial condition $\nu(x,0)$ described above. Equation (\ref{eq1}) is a semilinear parabolic PDE and is referred to as a scalar reaction-diffusion equation [@zheng2004nonlinear; @lawrenceevans2010]. To ensure existence and uniqueness of solutions to (\ref{eq1}), we further assume that for each $M>0$ there exists a constant $L_M>0$ depending on $M$ such that $\int_\mathbb{R}\nu_1(x)dx,\int_\mathbb{R}\nu_2(x)dx\leq M$ implies
\begin{equation}\label{local_lipschitz}
\int_\mathbb{R}\left|m(\nu_1,x)\nu_1(x)-m(\nu_2,x)\nu_2(x)\right|dx\leq L_M\int_\mathbb{R}|\nu_1(x)-\nu_2(x)|dx.
\end{equation}
In SM \S\ref{finite} we combine the above assumptions on growth rate, mutation and initial condition to prove that solutions to (\ref{eq1}) satisfy $N(t),|\bar x(t)|,\sigma^2(t)<+\infty$ for all $t\geq0$.

@@

Although clearly an idealized representation of biological reality, this model is sufficiently general to capture a large class of dynamics including density dependent growth and frequency dependent selection. As an example, logistic growth combined with quadratic stabilizing selection can be captured using the growth rate 
\begin{equation}\label{special_m}
m(\nu,x)=r-\frac{a}{2}(\theta-x)^2-c\int_{-\infty}^{+\infty}\nu(y,t)dy
\end{equation}
where $r\in\mathbb{R}$ is the intrinsic rate of growth in the absence of abiotic stabilizing selection, $a>0$ is the strength of abiotic stabilizing selection around the phenotypic optimum $\theta\in\mathbb{R}$, $c>0$ captures the sensitivity of fitness to intraspecific competition and $\kappa(x-y)=1$ so that competitive interactions cause the same reduction in fitness regardless of trait value. This exemplary fitness function has a few convenient properties. First, the effect of competition induces a local carrying capacity on the population, leading to a finite equilibrium abundance over bounded subsets of $\mathbb{R}$. Second, abiotic selection prevents the abundance density from diffusing too far from the abiotic optimum, leading to a finite equilibrial abundance across all of $\mathbb{R}$, $\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})$, finite equilibrial mean trait $\lim_{t\to\infty}\bar x(t)=\theta$ and finite equilibrial phenotypic variance $\lim_{t\to\infty}\sigma^2(t)=\sqrt{\frac{\mu}{a}}$, so long as $\bar x(0)\in\mathbb{R}$, $\sigma^2(0)\in[0,+\infty)$, $N(0)\in(0,+\infty)$ and $r>\frac{1}{2}\sqrt{a\mu}>0$. We prove these statements in SM \S\ref{equilib}. Note in particular the coincidence of equilibrial phenotypic variance predicted by this model and that predicted by the classic Gaussian allelic model of quantitative genetics [@Lande1975; @9780471986539; @Johnson2005]. In \S\ref{coev} and SM \S\ref{diffuse} we find this fitness function is closely related to a fitness function derived from niche theory that we use to model coevolution driven by resource competition.

@@

\begin{multline}\label{N_det}
\frac{d}{dt}N(t)=\frac{d}{dt}\int_{-\infty}^{+\infty}\nu(x,t)dx=\int_{-\infty}^{+\infty}\frac{\partial}{\partial t}\nu(x,t)dx \\
=\int_{-\infty}^{+\infty} m(\nu,x)\nu(x,t)dx+\int_{-\infty}^{+\infty}\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)dx \\
=N(t)\int_{-\infty}^{+\infty} m(\nu,x)p(x,t)dx=\bar m(t) N(t).
\end{multline}
Using our assumptions on mutation and rate of growth, we show in SM \S\ref{finite} $\nu(x,t)$ is twice differentiable with respect to $x$ and $\int_{-\infty}^{+\infty}\nu(x,t)dx<\infty$ for all $t\geq0$. This implies that we are justified in swapping the order of differentiation and integration and the result $\int_{-\infty}^{+\infty}\frac{\partial^2}{\partial x^2}\nu(x,t)dx=0$ can be derived from the fundamental theorem of calculus. Biological reasoning agrees with this latter result since mutation neither creates nor destroys individuals, but merely changes their type from their parental type. Taking the same approach, we derive the dynamics of the phenotypic distribution $p(x,t)$ in response to selection and muation as

\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=\Big(m(\nu,x)-\bar m(t)\Big) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}
This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which, for sexually reproducing populations, has been referred to as the Gaussian allelic approximation in reference to the distribution of mutational effects at loci in a genome on the values of traits [@Lande1975; but see also @9780471986539; along with @Johnson2005], the infinitesimal model in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect [@Barton2017] and the Gaussian descendants approximation in reference to offspring trait values being normally distributed around their parental values [@Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of Kimura's continuum-of-alleles model. We return to these concepts in \S\ref{inheritance} where we review a simple model of inheritance that leads us to an expression for the stochastic evolution of additive genetic variance. To distinguish this model from previous models of phenotypic evolution we refer to PDE (\ref{eq1}) from which (\ref{p_PDE}) was derived as the Deterministic Asexual Gaussian allelic model with Abundance dynamics (abbreviated DAGA). Later, we will extend this model to include the effects of demographic stochasticity, which we refer to as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA).

The covariance of fitness and phenotype across the population is defined as
\begin{equation}
\mathrm{Cov}_t\Big(m(\nu,x),x\Big)=\int_{-\infty}^{+\infty}\Big(m(\nu,x)-\bar m(t)\Big)\Big(x-\bar x(t)\Big)p(x,t)dx.
\end{equation}
Hence, the dynamics of the mean trait $\bar x(t)$ can be derived as

@@

Equation (\ref{detmean}) is a continuous time analog of the well known Robertson-Price equation without transmission bias [@Robertson1966; @PRICE1970; @FRANK2012; @Queller2017; @Lion2018]. Whether or not the covariance of fitness and phenotype creates change in $\bar x$ to maximize mean fitness $\bar m$ depends on the degree to which selection is frequency dependent [@Lande1976]. Since this change is driven by a covariance with respect to phenotypic diversity, the response in mean trait to selection is mediated by the phenotypic variance. In particular, when $\sigma^2=0$, $\bar x$ will not respond to selection. The result $\int_{-\infty}^{+\infty} x\frac{\partial^2}{\partial x^2} p(x,t)dx=0$ can be found by applying integration by parts. Following the approach taken to calculate the evolution of $\bar x$, we find the response of phenotypic variation to this model of selection and mutation is
\begin{equation}\label{vardyn_det}
\frac{d}{dt}\sigma^2(t)=\mathrm{Cov}_t\Big(m(\nu,x),(x-\bar x)^2\Big)+\mu.
\end{equation}
For the sake of space we relegate the derivation of $d\sigma^2/dt$ to SM \S\ref{var_deriv}. In the absence of mutation equation (\ref{vardyn_det}) agrees with the result derived by Lion [-@Lion2018] for discrete phenotypes. From a statistical perspective, if we think of $(x-\bar x)^2$ as a square error, then in analogy to the dynamics of the mean trait, we see that the response in $\sigma^2$ to selection can be expressed as a covariance of fitness and square error, which is defined in analogy to $\mathrm{Cov}_t(m(\nu,x),x)$. Just as for the evolution of $\bar x(t)$, this covariance also creates change in $\sigma^2$ that can either increase or decrease mean fitness $\bar m$, depending on whether or not individual fitness depends on the distribution of traits across the entire population. The effect of selection on phenotypic variance can be positive or negative depending on whether selection is stabilizing or disruptive.

@@

### Introduction to white noise

@@

```{r, echo=F, fig.align='center', fig.cap='\\label{wn}Approximations of sample paths of temporal white noise (left) and space-time white noise (right) with brightness scaled to value.'}
knitr::include_graphics('wn.png')
```

@@

We define a generalized stochastic process $\mathbf W$ that maps processes $f\in\mathscr{N}_2$ to real-valued stochastic processes indexed by time $t\geq0$, but not by space. To evaluate $\mathbf W$ for a process $f\in\mathscr{N}_2$ and some time $t\geq0$ we write $\mathbf W_t(f)$. Specifically, for any $f,g\in\mathscr{N}_2$, we define $\mathbf W(f)$ and $\mathbf W(g)$ to be Gaussian processes satisfying, for any $t,t_1,t_2\geq0$,
\begin{equation}\label{exp_WN}
\mathbb{E}\big(\mathbf W_t(f)\big)=\mathbb{E}\big(\mathbf W_t(g)\big)=0,
\end{equation}
\begin{equation}\label{cov_WN}
\mathbb{C}\big(\mathbf W_{t_1}(f),\mathbf W_{t_2}(g)\big)=\mathbb{E}\left(\int_0^{t_1\wedge t_2}\int_\mathbb{R} f(x,s)g(x,s)dxds\right),
\end{equation}
where $t_1\wedge t_2=\min(t_1,t_2)$ and $\mathbb{C}$ denotes covariance with respect to the underlying probability space. In particular, denoting $\mathbb{V}$ the variance operator with respect to the underlying probability space, we have $\mathbb{V}\big(\mathbf W_t(f)\big)=\|f\|_t^2$ for all $t\geq0$ and $f\in\mathscr{N}_2$. The operators $\mathbb{E}$ and $\mathbb{C}$ are to be distinguished from $\bar f(t)$ and $\mathrm{Cov}_t(f,g)$ which denote expectation and covariance with respect to phenotypic diversity at time $t\geq0$.

@@

# A model of diffuse coevolution \label{coev}

@@

__Ecological stoichiometry__

>>>>

---
title: A White Noise Approach to Evolutionary Ecology
author:
  - name: Bob Week
    email: bobweek@gmail.com
    affiliation: Bio
    footnote: 1
  - name: Steve Krone
    affiliation: Stats
  - name: Scott L. Nuismer
    affiliation: Bio
  - name: Luke J. Harmon
    affiliation: Bio
address:
  - code: Bio
    address: University of Idaho, Department of Biological Sciences, 875 Perimeter Drive MS 3051, Moscow, Idaho, 83844
  - code: Stats
    address: University of Idaho, Department of Statistical Science, 875 Perimeter Drive MS 1104, Moscow, Idaho, 83844
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued stochastic processes. This approach allows us to compute the full implications of demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a quantitative genetic model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests absolute values of linear selection gradients are negatively correlated with competition coefficients when the ratio of interspecific trait variation to intraspecific trait variation is approximately less than $1.25$ and positively correlated otherwise. The inverse relationship holds for quadratic selection gradients except with threshold diversity ratio equal to unity instead of $1.25$. Hence, this approach contributes to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of stochastic models that underlie rigorous investigations of the relationship between feedbacks of biological processes and the patterns of diversity they produce.
journal: "bioarxiv"
date: "`r Sys.Date()`"
bibliography: ref.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
output: 
  rticles::elsevier_article:
    extra_dependencies:
      - float
      - flafter
    fig_caption: yes
    fig_height: 3
    fig_width: 4
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{xr} # for cross-referencing between documents
  - \usepackage{mathtools}
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage{amsmath}
  - \externaldocument{SM} # externel document to cross-reference
  - \modulolinenumbers[2]
---

@@

Introduction
==========================

@@

The framework
============

@@

## Deterministic dynamics \label{deterministic}
Our review begins by considering the dynamics of an asexually reproducing population in a homogeneous environment. For simplicity, we first assume individuals are haploid and carry one of $K$ alleles each with a different fitness expressed as growth rate before introducing a model involving a quantitative trait. Under these assumptions, the derivation of the evolution of allele frequencies due to natural selection can be derived from expressions of exponential growth. This, and a few related approaches, have been provided by Crow and Kimura [-@9781932846126, \S 5.3]. Specifically, denoting $\nu_i$ the abundance of individuals with allele $i$ and $m_i$ the growth rate of allele $i$ [called the Malthusian parameter in @9781932846126], we have

\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation}

Starting from this model, we get the total abundance of the population as $N=\sum_{i=1}^K\nu_i$, the frequency of allele $i$ as $p_i=\nu_i/N$ and the mean fitness of the population as $\bar m=\sum_{i=1}^Kp_im_i$. Hence, we can employ some elementary calculus to derive the dynamics of abundance $dN/dt$ and the dynamics of allele frequencies $dp_1/dt,\dots,dp_K/dt$ as

\begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation}
\begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation}

Two important observations of these equations include (i) mean fitness $\bar m$ determines the abundance dynamics of the entire population and (ii) allele $i$ will increase (decrease) in frequency if $m_i>\bar m$ ($<\bar m$). Equation (\ref{popgen}) is known in the field of evolutionary game theory as the replicator equation [@Hofbauer1998; @9780674023383; @Lion2018; @Taylor1978; @Schuster1983]. Instead of being explicitly focused on alleles, the replicator equation describes the fluctuations of relative abundances of various _types_ in a population in terms of the vital rates of each type. Using a matrix of transition rates between differing types, it is straight-forward to extend the replicator equation to include mutation, which is known as the replicator-mutator equation [@9780674023383].

@@

To stay within the realm of biological plausibility we require a set technical assumptions. First, we assume the initial abundance density is continuous, non-negative, integrable and has finite trait mean and variance. That is, we assume $\nu(x,0)$ is continuous in $x$, satisfies $\nu(x,0)\geq0$ for all $x\in\mathbb{R}$ and

\begin{equation}
N(0)=\int_{-\infty}^{+\infty}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{-\infty}^{+\infty}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{-\infty}^{+\infty}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}

where $\bar x(t)$ and $\sigma^2(t)$ are respectively the mean trait and phenotypic variance at time $t\geq0$. Second, we assume selection is determined by the growth rate $m(h,x)$ that is differentiable with respect to both arguments and is bounded above by some $r\in\mathbb{R}$ for all $x\in\mathbb{R}$ and $h\geq0$. We also assume $(1+h)C\leq m(h,x)$ for some constant $C>0$ so that $m(h,x)$ is linear or more than linear in $h$. We denote by $K$ an operator that takes functions such as $\nu(x,t)$ as its argument and returns other functions that may still depend on the spatial variable $x$ and the temporal variable $t$. For our application we will be focused on operators $K$ that can be written as $(K\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x-y)\nu(y,t)dy$ for some non-negative and bounded function $\kappa$. We set $h=(K\nu)(x,t)$ to account for nonlocal effects of the abundance density $\nu(x,t)$ on the fitness of individuals with trait value $x$ [@Volpert2014]. As an example of a nonlocal effect, consider a population experiencing intraspecific competition for a common resource such that the competitive interactions are not mediated by any set of traits. In this case all individuals in the population effect the fitness of one another even though they may exhibit a diverse range of trait values. The term nonlocal comes from interpreting the trait value as spatial location. From here on we abbreviate $m(h,x)=m\big((K\nu)(x,t),x\big)$ to just $m(\nu,x)$. Third, we assume mutation is captured by diffusion with coefficient $\frac{\mu}{2}$. With these technicalities addressed we model the demographic dynamics of a population and the dynamics of a quantitative character simultaneously by the PDE

\begin{equation}\label{eq1}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)
\end{equation}

with the initial condition $\nu(x,0)$ described above. Equation (\ref{eq1}) is a semilinear parabolic PDE and is referred to as a scalar reaction-diffusion equation [@zheng2004nonlinear; @lawrenceevans2010]. To ensure existence and uniqueness of solutions to (\ref{eq1}), we further assume that for each $M>0$ there exists a constant $L_M>0$ depending on $M$ such that $\int_\mathbb{R}\nu_1(x)dx,\int_\mathbb{R}\nu_2(x)dx\leq M$ implies

\begin{equation}\label{local_lipschitz}
\int_\mathbb{R}\left|m(\nu_1,x)\nu_1(x)-m(\nu_2,x)\nu_2(x)\right|dx\leq L_M\int_\mathbb{R}|\nu_1(x)-\nu_2(x)|dx.
\end{equation}

In SM \S\ref{finite} we combine the above assumptions on growth rate, mutation and initial condition to prove that solutions to (\ref{eq1}) satisfy $N(t),|\bar x(t)|,\sigma^2(t)<+\infty$ for all $t\geq0$.

@@

Although clearly an idealized representation of biological reality, this model is sufficiently general to capture a large class of dynamics including density dependent growth and frequency dependent selection. As an example, logistic growth combined with quadratic stabilizing selection can be captured using the growth rate

\begin{equation}\label{special_m}
m(\nu,x)=r-\frac{a}{2}(\theta-x)^2-c\int_{-\infty}^{+\infty}\nu(y,t)dy
\end{equation}

where $r\in\mathbb{R}$ is the intrinsic rate of growth in the absence of abiotic stabilizing selection, $a>0$ is the strength of abiotic stabilizing selection around the phenotypic optimum $\theta\in\mathbb{R}$, $c>0$ captures the sensitivity of fitness to intraspecific competition and $\kappa(x-y)=1$ so that competitive interactions cause the same reduction in fitness regardless of trait value. This exemplary fitness function has a few convenient properties. First, the effect of competition induces a local carrying capacity on the population, leading to a finite equilibrium abundance over bounded subsets of $\mathbb{R}$. Second, abiotic selection prevents the abundance density from diffusing too far from the abiotic optimum, leading to a finite equilibrial abundance across all of $\mathbb{R}$, $\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})$, finite equilibrial mean trait $\lim_{t\to\infty}\bar x(t)=\theta$ and finite equilibrial phenotypic variance $\lim_{t\to\infty}\sigma^2(t)=\sqrt{\frac{\mu}{a}}$, so long as $\bar x(0)\in\mathbb{R}$, $\sigma^2(0)\in[0,+\infty)$, $N(0)\in(0,+\infty)$ and $r>\frac{1}{2}\sqrt{a\mu}>0$. We prove these statements in SM \S\ref{equilib}. Note in particular the coincidence of equilibrial phenotypic variance predicted by this model and that predicted by the classic Gaussian allelic model of quantitative genetics [@Lande1975; @9780471986539; @Johnson2005]. In \S\ref{coev} and SM \S\ref{diffuse} we find this fitness function is closely related to a fitness function derived from niche theory that we use to model coevolution driven by resource competition.

@@

\begin{multline}\label{N_det}
\frac{d}{dt}N(t)=\frac{d}{dt}\int_{-\infty}^{+\infty}\nu(x,t)dx=\int_{-\infty}^{+\infty}\frac{\partial}{\partial t}\nu(x,t)dx \\
=\int_{-\infty}^{+\infty} m(\nu,x)\nu(x,t)dx+\int_{-\infty}^{+\infty}\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)dx \\
=N(t)\int_{-\infty}^{+\infty} m(\nu,x)p(x,t)dx=\bar m(t) N(t).
\end{multline}

Using our assumptions on mutation and rate of growth, we show in SM \S\ref{finite} $\nu(x,t)$ is twice differentiable with respect to $x$ and $\int_{-\infty}^{+\infty}\nu(x,t)dx<\infty$ for all $t\geq0$. This implies that we are justified in swapping the order of differentiation and integration and the result $\int_{-\infty}^{+\infty}\frac{\partial^2}{\partial x^2}\nu(x,t)dx=0$ can be derived from the fundamental theorem of calculus. Biological reasoning agrees with this latter result since mutation neither creates nor destroys individuals, but merely changes their type from their parental type. Taking the same approach, we derive the dynamics of the phenotypic distribution $p(x,t)$ in response to selection and muation as

\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=\Big(m(\nu,x)-\bar m(t)\Big) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}

This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which, for sexually reproducing populations, has been referred to as the Gaussian allelic approximation in reference to the distribution of mutational effects at loci in a genome on the values of traits [@Lande1975; but see also @9780471986539; along with @Johnson2005], the infinitesimal model in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect [@Barton2017] and the Gaussian descendants approximation in reference to offspring trait values being normally distributed around their parental values [@Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of Kimura's continuum-of-alleles model. We return to these concepts in \S\ref{inheritance} where we review a simple model of inheritance that leads us to an expression for the stochastic evolution of additive genetic variance. To distinguish this model from previous models of phenotypic evolution we refer to PDE (\ref{eq1}) from which (\ref{p_PDE}) was derived as the Deterministic Asexual Gaussian allelic model with Abundance dynamics (abbreviated DAGA). Later, we will extend this model to include the effects of demographic stochasticity, which we refer to as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA).

The covariance of fitness and phenotype across the population is defined as

\begin{equation}
\mathrm{Cov}_t\Big(m(\nu,x),x\Big)=\int_{-\infty}^{+\infty}\Big(m(\nu,x)-\bar m(t)\Big)\Big(x-\bar x(t)\Big)p(x,t)dx.
\end{equation}

Hence, the dynamics of the mean trait $\bar x(t)$ can be derived as

@@

Equation (\ref{detmean}) is a continuous time analog of the well known Robertson-Price equation without transmission bias [@Robertson1966; @PRICE1970; @FRANK2012; @Queller2017; @Lion2018]. Whether or not the covariance of fitness and phenotype creates change in $\bar x$ to maximize mean fitness $\bar m$ depends on the degree to which selection is frequency dependent [@Lande1976]. Since this change is driven by a covariance with respect to phenotypic diversity, the response in mean trait to selection is mediated by the phenotypic variance. In particular, when $\sigma^2=0$, $\bar x$ will not respond to selection. The result $\int_{-\infty}^{+\infty} x\frac{\partial^2}{\partial x^2} p(x,t)dx=0$ can be found by applying integration by parts. Following the approach taken to calculate the evolution of $\bar x$, we find the response of phenotypic variation to this model of selection and mutation is

\begin{equation}\label{vardyn_det}
\frac{d}{dt}\sigma^2(t)=\mathrm{Cov}_t\Big(m(\nu,x),(x-\bar x)^2\Big)+\mu.
\end{equation}

For the sake of space we relegate the derivation of $d\sigma^2/dt$ to SM \S\ref{var_deriv}. In the absence of mutation equation (\ref{vardyn_det}) agrees with the result derived by Lion [-@Lion2018] for discrete phenotypes. From a statistical perspective, if we think of $(x-\bar x)^2$ as a square error, then in analogy to the dynamics of the mean trait, we see that the response in $\sigma^2$ to selection can be expressed as a covariance of fitness and square error, which is defined in analogy to $\mathrm{Cov}_t(m(\nu,x),x)$. Just as for the evolution of $\bar x(t)$, this covariance also creates change in $\sigma^2$ that can either increase or decrease mean fitness $\bar m$, depending on whether or not individual fitness depends on the distribution of traits across the entire population. The effect of selection on phenotypic variance can be positive or negative depending on whether selection is stabilizing or disruptive.

@@

@@

```{r, echo=F, fig.align='center', out.width='80%', fig.cap='\\label{wn}Approximations of sample paths of temporal white noise (left) and space-time white noise (right) with brightness scaled to value.'}
knitr::include_graphics('wn.png')
```

@@

We define a generalized stochastic process $\mathbf W$ that maps processes $f\in\mathscr{N}_2$ to real-valued stochastic processes indexed by time $t\geq0$, but not by space. To evaluate $\mathbf W$ for a process $f\in\mathscr{N}_2$ and some time $t\geq0$ we write $\mathbf W_t(f)$. Specifically, for any $f,g\in\mathscr{N}_2$, we define $\mathbf W(f)$ and $\mathbf W(g)$ to be Gaussian processes satisfying, for any $t,t_1,t_2\geq0$,

\begin{equation}\label{exp_WN}
\mathbb{E}\big(\mathbf W_t(f)\big)=\mathbb{E}\big(\mathbf W_t(g)\big)=0,
\end{equation}
\begin{equation}\label{cov_WN}
\mathbb{C}\big(\mathbf W_{t_1}(f),\mathbf W_{t_2}(g)\big)=\mathbb{E}\left(\int_0^{t_1\wedge t_2}\int_\mathbb{R} f(x,s)g(x,s)dxds\right),
\end{equation}

where $t_1\wedge t_2=\min(t_1,t_2)$ and $\mathbb{C}$ denotes covariance with respect to the underlying probability space. In particular, denoting $\mathbb{V}$ the variance operator with respect to the underlying probability space, we have $\mathbb{V}\big(\mathbf W_t(f)\big)=\|f\|_t^2$ for all $t\geq0$ and $f\in\mathscr{N}_2$. The operators $\mathbb{E}$ and $\mathbb{C}$ are to be distinguished from $\bar f(t)$ and $\mathrm{Cov}_t(f,g)$ which denote expectation and covariance with respect to phenotypic diversity at time $t\geq0$.

@@

A model of diffuse coevolution \label{coev}
============

@@

## Ecological stoichiometry
